\pdftitle

### 2018-05-15 maxNt_v3.tex
V2

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2018-05-15T01:35:24+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

One such quantitative criterion relies on the \me\ method
\cite{jaynes1957,jaynes1963,sivia1996_r2006,meadetal1984}. This criterion
has been used in neuroscience at least since the 1990s, applied to data
recorded from brain areas as diverse as retina and motor cortex
\cite{mackay1991,martignonetal1995,bohteetal2000,amarietal2003,schneidmanetal2006,shlensetal2006,mackeetal2009b,roudietal2009c,tkaciketal2009,gerwinnetal2009,mackeetal2011,mackeetal2011b,ganmoretal2011,granotatedgietal2013,tkaciketal2014b,moraetal2015,shimazakietal2015},
and it has been subjected to mathematical and conceptual scrutiny
\cite{tkaciketal2006,roudietal2009,roudietal2009b,barreiroetal2010b,mackeetal2013,rostamietal2016_r2017}.

@@

We hurry to say that the description just given is not \emph{the} way to
quantify and characterize cooperativity and functional grouping, but just
\emph{one} way. It can surely be criticized from many points of view. Yet,
it is quantitative and more precise than an undefined notion of
\enquote{cooperativity}. Two people who apply it to the same data will
obtain the same numbers. Other protocols can be based on the \me\ method,
for example protocols that take into account the activities or joint
activities of individual neurons.

The purpose of the present work is to show that there is a problem in this
general way of applying the \me\ method to the data recorded from our
sampled neurons.

Suppose that the neurons were recorded with electrodes covering an area of
some square millimetres. This recording is a \emph{sample} of the activity
of the population covered by the electrode; or we can consider it as a
sample of an even more extended brain area.

The characterization of the cooperativity of the recorded sample would have
little meaning if we did not expect its results to generalize to a larger
area, at least that covered by the electrodes. In other words, we expect
that the conclusions drawn with \me\ methods about the sampled neurons
should somehow generalize or extrapolate at least to larger areas from
which those neurons were sampled. We are thus assuming that the recorded
sample is a \emph{representative sample} of some larger area.

The problem is this. We can apply the \me\ method to the sample, as
described in the above protocol, to generate probability distributions for
the state of the sample. But, if our sample is representative of a larger
population, we can apply the \me\ method to the larger population, using
the constraints obtained from the sample, and generate a probability
distribution for the larger population; the distribution for the sample is
then obtained by marginalization. Either application seems to have some
commendable features. However, \emph{the distributions obtained from these
  two applications differ}. Which choice is most meaningful?

% LM: I'd like to avoid common beginnings like "Today blah blah is
% important" or "Blah blah is interesting/fundamental". They feel more like
% a mechanical ritual than a real tentative of connecting with the reader.
This note is mainly addressed to neuroscientists interested in \me\
methods, but we would be pleased if its discussion of the probability of
\enquote{sampling} were useful to neuroscientists that use other
statistical methods to study the physical and dynamical characteristics of
brain areas via neuronal recordings.

Recent electrophysiological techniques \cite{berenyietal2014} in fact allow
experimenters to record samples comprising even a couple hundreds neurons
from specific brain areas. From the observation of these samples we expect
to learn something about all other neurons in the same brain area. That is,
we assume that they are a \enquote{representative sample}. We do not
elaborate on the various important purposes of such recordings here, but
stress that these sample sizes can be considered \enquote{large} because
their statistical analysis requires considerable computational power.

These computational costs are one, probably not the earliest, of the many
reasons why \me\ methods have been introduced in neuroscience. It would be
useful to somehow compress the statistical wealth of large neuronal
recordings into few quantities, like sample moments for example. This
compression might also entail interesting biological or functional
properties of neuronal activity. The standard \me\ method
\cite{jaynes1957,sivia1996_r2006,meadetal1984} accomplishes this kind of
compression: it associates a unique probability distribution with few
experimental quantities. But this is only one of its uses. It is also used
for various information-theoretic purposes or to generate reference
probability distributions
\cite{bohteetal2000,schneidmanetal2006,shlensetal2006,roudietal2009b,mackeetal2011,tkaciketal2014b,shimazakietal2015}.
In all these uses the \me\ distribution is chosen as the \enquote{maximally
  noncommittal} one \cite{jaynes1963}. This adjective means little without
further technical characterizations. Different works give different
characterizations, but in this paper we will use the quoted expression as
an umbrella term for all of them. Our results will not depend on the
specific characterization of \enquote{noncommittal}.

In this note we show that we must face a dilemma if we want to apply the
\me\ method to a representative sample to find a maximally noncommittal
distribution.

The dilemma is this. We can apply the \me\ method to the sample, using a
specified set of experimental constraints, and generate a probability
distribution for the state of the sample. But our sample is representative
of a larger population, in which it is biologically and functionally
embedded. We can apply the \me\ method to the larger population, using the
same constraints, and generate a probability distribution for its larger
state, and then find the distribution for the sample by marginalization.
Either application seems to have some commendable features. However,
\emph{the distributions obtained from these two applications differ}. It
goes without saying that if one is \enquote{noncommittal}, the other must
be \enquote{committal}. Which choice is most meaningful?

In the final discussion we present some arguments in favour of one choice.

In the rest of the paper we mathematically formulate this dilemma. To this
purpose we also present some probability relations relevant to sampling.
The relations we present are well-known in survey sampling and in the
pedagogic problem of drawing from an urn without replacement, yet they are
somewhat hard to find explicitly written in the neuroscientific literature,
so they may be of interest on their own. We briefly discuss also two very
different meaning of \enquote{representative} that lead to similar initial
probability assignments for sampling.

\iffalse In the final discussion we present some arguments in favour of one choice
in our dilemma, but we do not mean that to be our final answer. This note,
and its dilemma,  also have the more general purpose of reminding our
readers the importance of asking \emph{what is the question we're trying to
  answer with probability theory?} and \emph{which assumptions do we want
  or need to make?} \fi

Our mathematical analysis pertains to neurons modelled as binary units at
one specific instant or short window in time. It is possible to similarly
discuss multi-state neuron models and population dynamics; but for
simplicity neurons are here assumed to be in a fixed, active \enquote{$1$}
or inactive \enquote{$0$} state; and evolution, change, time correlations,
and similar concepts do not concern us. We consider \me\ models that
constrain various kinds of sample or population averages; these models are
often called \enquote{homogeneous}. The final discussion touches upon
\enquote{inhomogeneous} models as well.

The notation in this note follows ISO and ANSI standards
\citep{iso1993,ieee1993,nist1995} but for the use of the comma \enquote{,}
to denote logical conjunction. Probability notation follows Jaynes
\citep{jaynes1994_r2003}. By \enquote{probability} we mean a degree of
belief which \enquote{would be agreed by all rational men if there were any
  rational men} \cite{good1966}.

>>>>

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2018-05-15T11:12:59+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

One such quantitative criterion relies on the \me\ method
\cite{jaynes1957,jaynes1963,hobsonetal1973,sivia1996_r2006,meadetal1984}.
This criterion has been used in neuroscience at least since the 1990s,
applied to data recorded from brain areas as diverse as retina and motor
cortex
\cite{mackay1991,martignonetal1995,bohteetal2000,amarietal2003,schneidmanetal2006,shlensetal2006,mackeetal2009b,roudietal2009c,tkaciketal2009,gerwinnetal2009,mackeetal2011,mackeetal2011b,ganmoretal2011,granotatedgietal2013,tkaciketal2014b,moraetal2015,shimazakietal2015},
and it has been subjected to mathematical and conceptual scrutiny
\cite{tkaciketal2006,roudietal2009,roudietal2009b,barreiroetal2010b,mackeetal2013,rostamietal2016_r2017}.

@@

We hurry to say that the description just given is not \emph{the} way to
quantify and characterize cooperativity and functional grouping, but just
\emph{one} way. It can surely be criticized from many points of view. Yet,
it is quantitative and more precise than an undefined notion of
\enquote{cooperativity}. Two people who apply it to the same data will
obtain the same numbers. Other protocols can be based on the \me\ method,
for example protocols that take into account the activities or joint
activities of individual neurons, or even protocols that take into account
time dependence.

The purpose of the present work is to show that there is a problem in the
way the \me\ method as sketched above, or more complex variations of it, is
applied to the activity of the recorded neurons. The problem stems from the
fact that the recorded neurons are a sample from a larger population.

Suppose that the neurons were recorded with electrodes covering an area of
some square millimetres. This recording is a sample of the activity of the
population covered by the electrode; we can consider it as a sample of an
even more extended brain area. The characterization of the cooperativity of
the recorded sample would have little meaning if we did not expect its
results to generalize to a larger area, at least that covered by the
electrodes. In other words, we expect that the conclusions drawn with the
\me\ methods about the sampled neurons should somehow generalize or
extrapolate at least to larger areas from which those neurons were sampled.
We are thus assuming that the recorded sample is a \emph{representative
  sample} of some larger area. Probability theory tells us how to make
inferences from a sample to the larger population from which it is sampled
\parentext{see references below}.

Here is the problem. We can apply the \me\ method to the sample, as
described in the above protocol, to generate probability distributions for
the state of the sample. But, if our sample is representative of a larger
population, we can apply the \me\ method to the larger population, using
the constraints obtained from the sample, and generate a probability
distribution for the larger population; the distribution for the sample is
then obtained by marginalization. Either application seems to have some
commendable features. However, \emph{the distributions obtained from these
  two applications differ}. Which choice is most meaningful?

In this paper we develop the second way of applying the \me, at the level
of the larger population, and show that its results differ from the
application at the sample level. We also consider the case where the size
of the larger population is unknown.

To apply the \me\ method to the larger, unsampled population, it is
necessary to use probability relations relevant to sampling
\cites{ghoshetal1997}[parts~I,VI]{freedmanetal1978_r2007}[\chap~8]{gelmanetal1995_r2014}[\chap~3]{jaynes1994_r2003}.
The relations we present are well-known in survey sampling and in the
pedagogic problem of drawing from an urn without replacement, yet they are
somewhat hard to find explicitly written in the neuroscientific literature,
so they may be of interest on their own.

\iffalse In the final discussion we present some arguments in favour of one choice
in our dilemma, but we do not mean that to be our final answer. This note,
and its dilemma,  also have the more general purpose of reminding our
readers the importance of asking \emph{what is the question we're trying to
  answer with probability theory?} and \emph{which assumptions do we want
  or need to make?}

Our mathematical analysis pertains to neurons modelled as binary units at
one specific instant or short window in time. It is possible to similarly
discuss multi-state neuron models and population dynamics; but for
simplicity neurons are here assumed to be in a fixed, active \enquote{$1$}
or inactive \enquote{$0$} state; and evolution, change, time correlations,
and similar concepts do not concern us. We consider \me\ models that
constrain various kinds of sample or population averages; these models are
often called \enquote{homogeneous}. The final discussion touches upon
\enquote{inhomogeneous} models as well.
\fi

The notation and terminology in this note follows ISO and ANSI standards
\citep{iso1993,ieee1993,nist1995,iso2006,iso2006b} but for the use of the
comma \enquote{,} to denote logical conjunction. Probability notation
follows Jaynes \citep{jaynes1994_r2003}. By \enquote{probability} we mean a
degree of belief which \enquote{would be agreed by all rational men if
  there were any rational men} \cite{good1966}.
